---
title: "DK1 day ahead electricity price"
format:
  dashboard:
    theme: [cosmo, theme.scss]
    embed-resources: true
    nav-buttons:
      - icon: github
        href: https://github.com/jacobazu/dk1-price-forecasting
    include-in-header:
      text: |
        <link rel="preconnect" href="https://fonts.googleapis.com">
        <link href="https://fonts.googleapis.com/css2?family=Plus+Jakarta+Sans:wght@400;500;600;700&display=swap" rel="stylesheet">
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(dplyr)
library(readr)
library(lubridate)
library(plotly)

Sys.setlocale("LC_TIME", "C")   # English day and month names, also on a Danish computer

orange <- "#f97316"
grey   <- "#6b7280"
ink    <- "#1f2328"

# actual prices in DK time
actual <- readRDS("data/processed/prices_dk1_hourly.rds") |>
  mutate(local = with_tz(time, "Europe/Copenhagen"),
         date  = as_date(local),
         hour  = hour(local)) |>
  group_by(date, hour) |>
  summarise(actual = mean(price), .groups = "drop")

# live forecasts
fc <- read_csv("forecasts/forecasts_live.csv", show_col_types = FALSE) |>
  mutate(date = as_date(date))

latest    <- max(fc$date)
tomorrow  <- fc |> filter(date == latest)
today_act <- actual |> filter(date == latest - 1)

# scored live forecasts (last 30 days)
scored <- fc |>
  inner_join(actual, by = c("date", "hour")) |>
  filter(date >= max(date) - 29)

eur  <- function(x) format(round(x, 1), nsmall = 1)

# difference tomorrow vs today for the first tile
diff_today <- mean(tomorrow$lear) - mean(today_act$actual)
diff_html  <- if (is.na(diff_today)) "" else
  sprintf('<span class="%s">%s%s</span> vs today',
          if (diff_today > 0) "kpi-up" else "kpi-down",
          if (diff_today > 0) "+" else "−", eur(abs(diff_today)))
hour_label <- function(h) sprintf("%02d:00", h)

# numbers for the tiles
cheap <- tomorrow |> slice_min(lear, n = 1, with_ties = FALSE)
peak  <- tomorrow |> slice_max(lear, n = 1, with_ties = FALSE)
has_scores <- nrow(scored) > 0
mae <- function(col) if (has_scores) eur(mean(abs(scored$actual - scored[[col]]))) else "–"

# Models: a button appears automatically for every model column in forecasts_live.csv.
# To add a model later, add its column name, label and colour here.
model_labels <- c(lear      = "LEAR",
                  lear_x    = "LEAR + exogenous",
                  xgb       = "XGBoost",
                  ensemble  = "Ensemble",
                  naive_day = "Naive")
model_colors <- c(lear      = "#f97316",
                  lear_x    = "#0ea5e9",
                  xgb       = "#16a34a",
                  ensemble  = "#8b5cf6",
                  naive_day = "#9ca3af")
models <- intersect(names(model_labels), names(fc))

# buttons above a chart: one per model plus "All", actual prices are always shown
# html buttons instead of plotly's own, so they can be styled in theme.scss
model_buttons <- function(p, models) {
  vis <- c(lapply(models, \(m) c(TRUE, models == m)), list(rep(TRUE, length(models) + 1)))
  htmlwidgets::onRender(p, "
    function(el, x, data) {
      el.style.position = 'relative';
      var bar = document.createElement('div');
      bar.className = 'model-buttons';
      data.labels.forEach(function(lab, i) {
        var b = document.createElement('button');
        b.textContent = lab;
        if (i === 0) b.className = 'active';
        b.onclick = function() {
          Plotly.restyle(el, {visible: data.visible[i]});
          bar.querySelectorAll('button').forEach(function(o) { o.className = ''; });
          b.className = 'active';
        };
        bar.appendChild(b);
      });
      el.appendChild(bar);
    }",
    data = list(labels = c(unname(model_labels[models]), "All"), visible = vis))
}

chart_layout <- function(p, models, xtitle = "") {
  p |>
    layout(paper_bgcolor = "rgba(0,0,0,0)",
           plot_bgcolor  = "rgba(0,0,0,0)",
           font = list(family = "Segoe UI, Helvetica Neue, Arial, sans-serif", color = grey),
           xaxis = list(title = xtitle, showgrid = FALSE, zeroline = FALSE, color = grey,
                        linecolor = "#e5ddd3"),
           yaxis = list(title = "", color = grey, gridcolor = "#f1ebe4",
                        zerolinecolor = "#cfc6bb"),
           legend = list(orientation = "h", x = 0, y = -0.15, font = list(color = ink)),
           hovermode = "x unified",
           hoverlabel = list(bgcolor = "#ffffff", bordercolor = "#efe6dc",
                             font = list(color = ink)),
           margin = list(t = 50)) |>
    config(displayModeBar = FALSE) |>
    model_buttons(models)
}
```

# Forecast

## Row {height=14%}

::: {.card}
<div class="kpi" style="--accent: #f97316">
<div>
<div class="kpi-label">Average price tomorrow</div>
<div class="kpi-value">`r eur(mean(tomorrow$lear))`<span class="kpi-unit">€/MWh</span></div>
<div class="kpi-sub">`r diff_html`</div>
</div>
<div class="kpi-icon"><i class="bi bi-lightning-charge-fill"></i></div>
</div>
:::

::: {.card}
<div class="kpi" style="--accent: #16a34a">
<div>
<div class="kpi-label">Cheapest hour</div>
<div class="kpi-value">`r hour_label(cheap$hour)`</div>
<div class="kpi-sub">`r eur(cheap$lear)` €/MWh</div>
</div>
<div class="kpi-icon"><i class="bi bi-arrow-down-circle-fill"></i></div>
</div>
:::

::: {.card}
<div class="kpi" style="--accent: #dc2626">
<div>
<div class="kpi-label">Most expensive hour</div>
<div class="kpi-value">`r hour_label(peak$hour)`</div>
<div class="kpi-sub">`r eur(peak$lear)` €/MWh</div>
</div>
<div class="kpi-icon"><i class="bi bi-arrow-up-circle-fill"></i></div>
</div>
:::

::: {.card}
<div class="kpi" style="--accent: #c2410c">
<div>
<div class="kpi-label">Backtest accuracy</div>
<div class="kpi-value">−26%</div>
<div class="kpi-sub">error vs naive (MAE)</div>
</div>
<div class="kpi-icon"><i class="bi bi-bullseye"></i></div>
</div>
:::

## Row {height=86%}

```{r}
#| title: "Tomorrow's forecast vs today's prices (EUR/MWh)"
p <- plot_ly() |>
  add_lines(data = today_act, x = ~hour, y = ~actual,
            name = paste("Actual,", format(latest - 1, "%a %d %b")),
            line = list(color = grey, width = 2, dash = "dot"),
            hovertemplate = "%{y:.1f} €/MWh<extra>Actual</extra>")

# Naive is left out here: its forecast for tomorrow is simply today's prices (the dotted line)
fc_models <- setdiff(models, "naive_day")

for (m in fc_models) {
  p <- p |>
    add_lines(data = tomorrow, x = ~hour, y = tomorrow[[m]],
              name = paste0(model_labels[[m]], ", ", format(latest, "%a %d %b")),
              visible = (m == fc_models[1]),
              line = list(color = model_colors[[m]], width = 3),
              hovertemplate = paste0("%{y:.1f} €/MWh<extra>", model_labels[[m]], "</extra>"))
}

p |> chart_layout(fc_models, xtitle = "Hour (Danish time)") |> layout(xaxis = list(dtick = 3))
```

# Track record

## Row {height=14%}

::: {.card}
<div class="kpi" style="--accent: #f97316">
<div>
<div class="kpi-label">Days tracked</div>
<div class="kpi-value">`r n_distinct(scored$date)`</div>
<div class="kpi-sub">`r if (has_scores) paste("since", format(min(scored$date), "%d %b %Y")) else "starts after the first forecast day"`</div>
</div>
<div class="kpi-icon"><i class="bi bi-calendar-check-fill"></i></div>
</div>
:::

::: {.card}
<div class="kpi" style="--accent: #16a34a">
<div>
<div class="kpi-label">LEAR error (MAE)</div>
<div class="kpi-value">`r mae("lear")`</div>
<div class="kpi-sub">€/MWh, last 30 days</div>
</div>
<div class="kpi-icon"><i class="bi bi-graph-up-arrow"></i></div>
</div>
:::

::: {.card}
<div class="kpi" style="--accent: #6b7280">
<div>
<div class="kpi-label">Naive error (MAE)</div>
<div class="kpi-value">`r mae("naive_day")`</div>
<div class="kpi-sub">€/MWh, last 30 days</div>
</div>
<div class="kpi-icon"><i class="bi bi-arrow-repeat"></i></div>
</div>
:::

## Row {height=56%}

```{r}
#| title: "Live forecasts vs actual prices, last 30 days (EUR/MWh)"
if (nrow(scored) == 0) {
  cat("The track record starts once the first forecast day has passed.")
} else {
  sc <- scored |> mutate(time = as_datetime(date, tz = "Europe/Copenhagen") + hours(hour))

  p2 <- plot_ly(sc, x = ~time) |>
    add_lines(y = ~actual, name = "Actual",
              line = list(color = ink, width = 1.5),
              hovertemplate = "%{y:.1f} €/MWh<extra>Actual</extra>")

  for (m in models) {
    p2 <- p2 |>
      add_lines(y = sc[[m]], name = model_labels[[m]],
                visible = (m == models[1]),
                line = list(color = model_colors[[m]], width = 1.5),
                hovertemplate = paste0("%{y:.1f} €/MWh<extra>", model_labels[[m]], "</extra>"))
  }

  p2 |> chart_layout(models)
}
```

## Row {height=30%}

```{r}
#| title: "Accuracy by model, last 30 days"
if (nrow(scored) > 0) {
  lapply(models, \(m) {
    s <- scored[!is.na(scored[[m]]), ]
    tibble(Model            = model_labels[[m]],
           `Days tracked`   = n_distinct(s$date),
           `MAE (€/MWh)`    = mean(abs(s$actual - s[[m]])),
           `RMSE (€/MWh)`   = sqrt(mean((s$actual - s[[m]])^2)))
  }) |>
    bind_rows() |>
    arrange(`MAE (€/MWh)`) |>
    knitr::kable(digits = 1)
}
```

# Method

## Row

::: {.card title="Backtest: October 2025 – October 2026"}

Rolling out-of-sample forecasts for every day of the test year (≈ 8,800 hourly forecasts).

| Model | MAE (€/MWh) | RMSE (€/MWh) |
|---|---|---|
| **LEAR** | **22.7** | **35.8** |
| Naive: same hour yesterday | 30.8 | 46.8 |
| Naive: weekly mix | 34.2 | 51.4 |

LEAR cuts the average error by about 26% compared with the best naive benchmark.

:::

::: {.card title="How it works"}

**LEAR** inspired by Lago (2021): one LASSO
regression per hour, 24 regressions in total, each using all 24 prices from days D-1, D-2, D-3
and D-7 plus weekday dummies. Each forecast is estimated on the previous two years, with lambda
chosen by BIC. Prices are transformed with asinh((p − median) / MAD) to handle spikes and
negative prices.

**Live track record:** every morning before the noon auction, the forecast for the next day is
saved and never changed afterwards.

**Data:** [Energi Data Service](https://www.energidataservice.dk) (Energinet).<br>
**Code:** [github.com/jacobazu/dk1-price-forecasting](https://github.com/jacobazu/dk1-price-forecasting)<br>
**Last updated:** `r format(with_tz(Sys.time(), "Europe/Copenhagen"), "%d %B %Y, %H:%M")` (Danish time)

:::
