---
title: "DK1 Day-Ahead Electricity Price Forecast"
format:
  dashboard:
    theme: flatly
    embed-resources: true
    nav-buttons:
      - icon: github
        href: https://github.com/jacobazu/dk1-price-forecasting
    include-in-header:
      text: |
        <style>
          /* key-number tiles: white card, coloured strip on the left */
          .kpi { height: 100%; display: flex; flex-direction: column; justify-content: center;
                 padding: 0.2rem 0.9rem; border-left: 4px solid var(--accent, #1f5fbf); }
          .kpi-label { font-size: 0.8rem; color: #6c757d; text-transform: uppercase;
                       letter-spacing: 0.05em; }
          .kpi-value { font-size: 2.1rem; font-weight: 600; color: #2c3e50; line-height: 1.15; }
          .kpi-sub   { font-size: 0.9rem; color: #6c757d; }
        </style>
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(dplyr)
library(readr)
library(lubridate)
library(plotly)

Sys.setlocale("LC_TIME", "C")   # English day and month names, also on a Danish computer

blue <- "#1f5fbf"
grey <- "#9aa5b1"

# actual prices in DK time
actual <- readRDS("data/processed/prices_dk1_hourly.rds") |>
  mutate(local = with_tz(time, "Europe/Copenhagen"),
         date  = as_date(local),
         hour  = hour(local)) |>
  group_by(date, hour) |>
  summarise(actual = mean(price), .groups = "drop")

# live forecasts
fc <- read_csv("forecasts/forecasts_live.csv", show_col_types = FALSE) |>
  mutate(date = as_date(date))

latest    <- max(fc$date)
tomorrow  <- fc |> filter(date == latest)
today_act <- actual |> filter(date == latest - 1)

# scored live forecasts (last 30 days)
scored <- fc |>
  inner_join(actual, by = c("date", "hour")) |>
  filter(date >= max(date) - 29)

eur  <- function(x) format(round(x, 1), nsmall = 1)
hour_label <- function(h) sprintf("%02d:00", h)

# numbers for the tiles
cheap <- tomorrow |> slice_min(lear, n = 1, with_ties = FALSE)
peak  <- tomorrow |> slice_max(lear, n = 1, with_ties = FALSE)
has_scores <- nrow(scored) > 0
mae <- function(col) if (has_scores) eur(mean(abs(scored$actual - scored[[col]]))) else "–"

# Models: a button appears automatically for every model column in forecasts_live.csv.
# To add a model later, add its column name, label and colour here.
model_labels <- c(lear      = "LEAR",
                  lear_x    = "LEAR + exogenous",
                  xgb       = "XGBoost",
                  ensemble  = "Ensemble",
                  naive_day = "Naive")
model_colors <- c(lear      = "#1f5fbf",
                  lear_x    = "#e07b00",
                  xgb       = "#2a9d55",
                  ensemble  = "#7b3fbf",
                  naive_day = "#9aa5b1")
models <- intersect(names(model_labels), names(fc))

# Buttons above a chart: one per model, plus "All".
# The chart's first trace (actual prices) is always shown.
model_menu <- function(models) {
  n <- length(models)
  buttons <- lapply(models, \(m) list(
    method = "restyle",
    label  = model_labels[[m]],
    args   = list("visible", as.list(c(TRUE, models == m)))
  ))
  buttons <- c(buttons, list(list(method = "restyle", label = "All",
                                  args = list("visible", as.list(rep(TRUE, n + 1))))))
  list(list(type = "buttons", direction = "right", active = 0, showactive = TRUE,
            x = 0, xanchor = "left", y = 1.02, yanchor = "bottom",
            bgcolor = "#f4f6f9", bordercolor = "#d5dae1", font = list(size = 13),
            buttons = buttons))
}

chart_layout <- function(p, models, xtitle = "") {
  p |>
    layout(xaxis = list(title = xtitle, showgrid = FALSE),
           yaxis = list(title = "", gridcolor = "#eceff3", zerolinecolor = "#c8ced6"),
           legend = list(orientation = "h", x = 0, y = -0.15),
           hovermode = "x unified",
           margin = list(t = 50),
           updatemenus = model_menu(models)) |>
    config(displayModeBar = FALSE)
}
```

# Forecast

## Row {height=18%}

::: {.card}
<div class="kpi" style="--accent: #1f5fbf">
<div class="kpi-label">Average price tomorrow</div>
<div class="kpi-value">`r eur(mean(tomorrow$lear))`</div>
<div class="kpi-sub">€/MWh · today `r eur(mean(today_act$actual))`</div>
</div>
:::

::: {.card}
<div class="kpi" style="--accent: #2a9d55">
<div class="kpi-label">Cheapest hour</div>
<div class="kpi-value">`r hour_label(cheap$hour)`</div>
<div class="kpi-sub">`r eur(cheap$lear)` €/MWh</div>
</div>
:::

::: {.card}
<div class="kpi" style="--accent: #d64550">
<div class="kpi-label">Most expensive hour</div>
<div class="kpi-value">`r hour_label(peak$hour)`</div>
<div class="kpi-sub">`r eur(peak$lear)` €/MWh</div>
</div>
:::

::: {.card}
<div class="kpi" style="--accent: #9aa5b1">
<div class="kpi-label">Backtest accuracy</div>
<div class="kpi-value">−26%</div>
<div class="kpi-sub">error vs naive benchmark (MAE)</div>
</div>
:::

## Row

```{r}
#| title: "Tomorrow's forecast vs today's prices (EUR/MWh)"
p <- plot_ly() |>
  add_lines(data = today_act, x = ~hour, y = ~actual,
            name = paste("Actual,", format(latest - 1, "%a %d %b")),
            line = list(color = grey, width = 2, dash = "dot"),
            hovertemplate = "%{y:.1f} €/MWh<extra>Actual</extra>")

# Naive is left out here: its forecast for tomorrow is simply today's prices (the dotted line)
fc_models <- setdiff(models, "naive_day")

for (m in fc_models) {
  p <- p |>
    add_lines(data = tomorrow, x = ~hour, y = tomorrow[[m]],
              name = paste0(model_labels[[m]], ", ", format(latest, "%a %d %b")),
              visible = (m == fc_models[1]),
              line = list(color = model_colors[[m]], width = 3),
              hovertemplate = paste0("%{y:.1f} €/MWh<extra>", model_labels[[m]], "</extra>"))
}

p |> chart_layout(fc_models, xtitle = "Hour (Danish time)") |> layout(xaxis = list(dtick = 3))
```

# Track record

## Row {height=18%}

::: {.card}
<div class="kpi" style="--accent: #1f5fbf">
<div class="kpi-label">Days tracked</div>
<div class="kpi-value">`r n_distinct(scored$date)`</div>
<div class="kpi-sub">`r if (has_scores) paste("since", format(min(scored$date), "%d %b %Y")) else "starts after the first forecast day"`</div>
</div>
:::

::: {.card}
<div class="kpi" style="--accent: #2a9d55">
<div class="kpi-label">LEAR error (MAE)</div>
<div class="kpi-value">`r mae("lear")`</div>
<div class="kpi-sub">€/MWh, last 30 days</div>
</div>
:::

::: {.card}
<div class="kpi" style="--accent: #9aa5b1">
<div class="kpi-label">Naive error (MAE)</div>
<div class="kpi-value">`r mae("naive_day")`</div>
<div class="kpi-sub">€/MWh, last 30 days</div>
</div>
:::

## Row

```{r}
#| title: "Live forecasts vs actual prices, last 30 days (EUR/MWh)"
if (nrow(scored) == 0) {
  cat("The track record starts once the first forecast day has passed.")
} else {
  sc <- scored |> mutate(time = as_datetime(date, tz = "Europe/Copenhagen") + hours(hour))

  p2 <- plot_ly(sc, x = ~time) |>
    add_lines(y = ~actual, name = "Actual",
              line = list(color = "#2b2d42", width = 1.5),
              hovertemplate = "%{y:.1f} €/MWh<extra>Actual</extra>")

  for (m in models) {
    p2 <- p2 |>
      add_lines(y = sc[[m]], name = model_labels[[m]],
                visible = (m == models[1]),
                line = list(color = model_colors[[m]], width = 1.5),
                hovertemplate = paste0("%{y:.1f} €/MWh<extra>", model_labels[[m]], "</extra>"))
  }

  p2 |> chart_layout(models)
}
```

## Row {height=30%}

```{r}
#| title: "Accuracy by model, last 30 days"
if (nrow(scored) > 0) {
  lapply(models, \(m) {
    s <- scored[!is.na(scored[[m]]), ]
    tibble(Model            = model_labels[[m]],
           `Days tracked`   = n_distinct(s$date),
           `MAE (€/MWh)`    = mean(abs(s$actual - s[[m]])),
           `RMSE (€/MWh)`   = sqrt(mean((s$actual - s[[m]])^2)))
  }) |>
    bind_rows() |>
    arrange(`MAE (€/MWh)`) |>
    knitr::kable(digits = 1)
}
```

# Method

## Row

::: {.card title="Backtest: October 2025 – October 2026"}

Rolling out-of-sample forecasts for every day of the test year (≈ 8,800 hourly forecasts).

| Model | MAE (€/MWh) | RMSE (€/MWh) |
|---|---|---|
| **LEAR** | **22.7** | **35.8** |
| Naive: same hour yesterday | 30.8 | 46.8 |
| Naive: weekly mix | 34.2 | 51.4 |

LEAR cuts the average error by about 26% compared with the best naive benchmark.

:::

::: {.card title="How it works"}

**LEAR** (LASSO-estimated autoregressive model, inspired by Lago (2021): one LASSO
regression per hour, 24 regressions in total, each using all 24 prices from days D-1, D-2, D-3
and D-7 plus weekday dummies. Each forecast is estimated on the previous two years, with lambda
chosen by BIC. Prices are transformed with asinh((p − median) / MAD) to handle spikes and
negative prices.

**Live track record:** every morning before the noon auction, the forecast for the next day is
saved and never changed afterwards.

**Data:** [Energi Data Service](https://www.energidataservice.dk) (Energinet).<br>
**Code:** [github.com/jacobazu/dk1-price-forecasting](https://github.com/jacobazu/dk1-price-forecasting)<br>
**Last updated:** `r format(with_tz(Sys.time(), "Europe/Copenhagen"), "%d %B %Y, %H:%M")` (Danish time)

:::
