---
title: "DK1 day ahead electricity price forecast"
subtitle: "Denmark (west) · before noon auction, daily update"
format:
  html:
    embed-resources: true
    theme: cosmo
    page-layout: article
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(readr)
library(lubridate)
library(ggplot2)
library(plotly)

# actual prices in DK time
actual <- readRDS("data/processed/prices_dk1_hourly.rds") |>
  mutate(local = with_tz(time, "Europe/Copenhagen"),
         date  = as_date(local),
         hour  = hour(local)) |>
  group_by(date, hour) |>
  summarise(actual = mean(price), .groups = "drop")

# live forecasts
fc <- read_csv("forecasts/forecasts_live.csv", show_col_types = FALSE) |>
  mutate(date = as_date(date))

latest <- max(fc$date)
```

## Forecast for `r format(latest, "%A %d %B %Y")`

```{r}
plot_df <- bind_rows(
  actual |> filter(date == latest - 1) |>
    transmute(hour, price = actual, series = paste("Actual,", format(latest - 1, "%d %b"))),
  fc |> filter(date == latest) |>
    transmute(hour, price = lear, series = paste("LEAR forecast,", format(latest, "%d %b")))
)

p <- ggplot(plot_df, aes(hour, price, colour = series, linetype = series)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("grey50", "#1f5fbf")) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(x = "Hour (Danish time)", y = "EUR/MWh", colour = NULL, linetype = NULL) +
  theme_minimal()

ggplotly(p) |> layout(legend = list(orientation = "h", y = -0.2))
```

## Live track record

Every forecast is saved before the auction and never changed afterwards.

```{r}
scored <- fc |>
  inner_join(actual, by = c("date", "hour")) |>
  filter(date >= max(date) - 29)

if (nrow(scored) == 0) {
  cat("The track record starts once the first forecast day has passed.")
} else {
  p2 <- scored |>
    mutate(time = as_datetime(date, tz = "Europe/Copenhagen") + hours(hour)) |>
    ggplot(aes(time)) +
    geom_line(aes(y = actual, colour = "Actual"), linewidth = 0.5) +
    geom_line(aes(y = lear,   colour = "LEAR forecast"), linewidth = 0.5) +
    scale_colour_manual(values = c("Actual" = "black", "LEAR forecast" = "#1f5fbf")) +
    labs(x = NULL, y = "EUR/MWh", colour = NULL) +
    theme_minimal()

  ggplotly(p2) |> layout(legend = list(orientation = "h", y = -0.2))
}
```

```{r}
if (nrow(scored) > 0) {
  tibble(
    Model = c("LEAR", "Naive (same hour yesterday)"),
    `MAE (EUR/MWh)`  = c(mean(abs(scored$actual - scored$lear)),
                         mean(abs(scored$actual - scored$naive_day))),
    `RMSE (EUR/MWh)` = c(sqrt(mean((scored$actual - scored$lear)^2)),
                         sqrt(mean((scored$actual - scored$naive_day)^2)))
  ) |>
    knitr::kable(digits = 1, caption = paste("Last", n_distinct(scored$date), "forecast days"))
}
```

## Backtest (Oct 2025 – Oct 2026)

| Model | MAE | RMSE |
|---|---|---|
| **LEAR** | **22.7** | **35.8** |
| Naive: same hour yesterday | 30.8 | 46.8 |
| Naive: weekly mix | 34.2 | 51.4 |

## Method

**LEAR** (LASSO-estimated autoregressive model, Lago et al. 2021): one LASSO regression per hour,
using all 24 prices from day D-1, D-2, D-3 and D-7 plus weekday dummies. Each forecast is
estimated on the previous two years, with lambda chosen by BIC. Prices are transformed with
asinh((p − median) / MAD) to handle spikes and negative prices.

Data: [Energi Data Service](https://www.energidataservice.dk) (Energinet).
Code: [github.com/jacobazu/dk1-price-forecasting](https://github.com/jacobazu/dk1-price-forecasting).
Last updated: `r format(with_tz(Sys.time(), "Europe/Copenhagen"), "%d %B %Y, %H:%M")` (Danish time).
